---
title: "LPAIV-Infected and Controls by DPI - Bursa - Gene"
author: "Amanda Dolinski & Jared J. Homola"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

\pagenumbering{gobble}

This is an analysis of differential expression at the gene level between LPAIV-infected and uninfected control blue-winged teal bursa samples at each day post-infection (DPI). Differential expression was also analyzed between LPAIV-infected groups at different DPI.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.pos = 'H')
```

```{r message=FALSE, warning=FALSE}
## Load packages and data
library(limma)
library(edgeR)
library(Glimma)
library(gplots)
library(ggrepel)
library(RColorBrewer)
library(gridExtra)
library(kableExtra)
library(topGO)
library(tidyverse)

## Load data
annot <- read.delim("G:/Shared drives/RNA Seq Supershedder Project/BWTE DE manuscript/extData/Trinotate.csv", header = TRUE, sep = "\t")
cnt <- read.table("G:/Shared drives/RNA Seq Supershedder Project/BWTE DE manuscript/extData/rsem.gene.counts.matrix", header = TRUE)
covars <- read.csv("G:/Shared drives/RNA Seq Supershedder Project/BWTE DE manuscript/extData/BWTE54_SSgroup_Raw_Pool.csv", header = TRUE)
```

```{r, warning = FALSE, message = FALSE}
cnt <- cnt %>%
  select(
    -alignEstimateAbundance_BWTE_Ileum_36_S50,
    -alignEstimateAbundance_BWTE_Bursa_36_S31,-alignEstimateAbundance_BWTE_Ileum_19_S35,
    -alignEstimateAbundance_BWTE_Bursa_19_S14) %>%
  rownames_to_column("gene") %>%
  separate(gene, into = c(NA, "pt1", "pt2", "gene")) %>%
  unite(gene, pt1, pt2, gene) %>%
  column_to_rownames("gene") %>% 
  select(contains("Bursa"))

covars <- covars %>%
  filter(!bird %in% c("36", "19")) %>%
  arrange(bird) %>%
  mutate(group = str_remove(group, "-"))

annot <- annot %>%
  separate(transcript_id, into = c(NA, "pt1", "pt2", "gene", "isoform")) %>%
  unite(transcript_id, pt1, pt2, gene, isoform) %>% 
  separate(gene_id, into = c(NA, "pt1", "pt2", "gene")) %>%
  unite(gene_id, pt1, pt2, gene)
```

```{r, warning = FALSE, message = FALSE}
bird <- as.factor(covars$bird)
sex <- as.factor(covars$sex)
age <- as.numeric(covars$age)
weight <- covars$wt_55
group <- recode(covars$group, C1 = "Ctl", C14 = "Ctl") %>% 
  as.factor()
pool <- as.factor(covars$Pool.Bursa)
```

```{r, warning = FALSE, message = FALSE, results = 'hide'}
#Convert to DGEList object
dge <- DGEList(counts=cnt)
dge$genes <- annot[match(rownames(dge$counts), annot$gene_id),]

#CPM and log-CPM
cpm <- cpm(dge)
lcpm <- cpm(dge, log = TRUE)

#### Retain only genes expressed at >0.5 CPM in at least 25% of individuals
table(rowSums(dge$counts==0) == length(cnt[1,]))
keep.exprs <- rowSums(cpm>0.5) >= length(cnt[1,])/4
sum(keep.exprs)

dge <- dge[keep.exprs,, keep.lib.sizes=FALSE]
table(rowSums(dge$counts==0) == length(cnt[1,]))
dim(dge)

#TMM normalization
dge <- calcNormFactors(dge, method = "TMM")
```


\newpage 

# Differential expression analysis
```{r}
## Establish design matrix
design = model.matrix(~ 0 +
                        group +
                        age +
                        sex + 
                        weight + 
                        pool)

colnames(design) <- gsub("group", "", colnames(design))

contr.matrix = makeContrasts(
  CtlvI1 = Ctl - I1,
  CtlvI3 = Ctl - I3,
  CtlvI5 = Ctl - I5,
  CtlvI14 = Ctl - I14,
  I1vI3 = I1 - I3,
  I3vI5 = I3 - I5,
  I5vI14 = I5 - I14,
  levels = colnames(design))

v <- voomWithQualityWeights(dge, design, plot = FALSE)

vfit <- lmFit(v, design) 
vfit <- contrasts.fit(vfit, contrasts = contr.matrix)
tfit <- treat(vfit, lfc = 1.0)
```

```{r}
dt <- decideTests(tfit, p.value = 0.1, adjust.method = "fdr")
summary(dt) %>%
  kbl(booktabs = T, caption = "Count of DE genes. For a gene to be considered differentially expressed, we require a p-value of 0.1 with a false discovery rate correction and a log fold change difference of 1.0.") %>%
  kable_styling(position = "center",
                latex_options = c("hold_position"))

dt.tib <- as_tibble(dt, rownames = NA) %>% 
  rownames_to_column("gene") %>% 
  mutate_at(vars(starts_with("C")), as.numeric) %>% 
  mutate_at(vars(starts_with("I")), as.numeric) %>% 
  filter_at(vars(CtlvI1:I5vI14), any_vars(. != 0))
```

\newpage 

### Volcano plot  
Volcano plot reporting -log10(p-values) as a function of log2(fold change) between the samples (logFC, x axis). Genes that are identified as significantly differentially expressed following a false discovery rate correction (q = 0.10) are shown in red 
```{r, warning = FALSE, message = FALSE, fig.height = 8, fig.width = 7, fig.align = "center"}
tmp1 <- topTreat(tfit, coef = 1, n = Inf)
results1 <- mutate(tmp1, sig=ifelse(tmp1$adj.P.Val<0.1, "Sig", "Not Sig"))
p1 <- ggplot(results1, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Control vs. Infected, Day 1") +
  ylab("-Log10(Adjusted p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none",
        plot.title = element_text(size = 9))

tmp2 <- topTreat(tfit, coef = 2, n = Inf)
results2 <- mutate(tmp2, sig=ifelse(tmp2$adj.P.Val<0.1, "Sig", "Not Sig"))
p2 <- ggplot(results2, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Control vs. Infected, Day 3") +
  ylab("-Log10(Adjusted p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none",
        plot.title = element_text(size = 9))

tmp3 <- topTreat(tfit, coef = 3, n = Inf)
results3 <- mutate(tmp3, sig=ifelse(tmp3$adj.P.Val<0.1, "Sig", "Not Sig"))
p3 <- ggplot(results3, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Control vs. Infected, Day 5") +
  ylab("-Log10(Adjusted p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none",
        plot.title = element_text(size = 9))

tmp4 <- topTreat(tfit, coef = 4, n = Inf)
results4 <- mutate(tmp4, sig=ifelse(tmp4$adj.P.Val<0.1, "Sig", "Not Sig"))
p4 <- ggplot(results4, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Control vs. Infected, Day 14") +
  ylab("-Log10(Adjusted p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none",
        plot.title = element_text(size = 9))

tmp5 <- topTreat(tfit, coef = 5, n = Inf)
results5 <- mutate(tmp5, sig=ifelse(tmp5$adj.P.Val<0.1, "Sig", "Not Sig"))
p5 <- ggplot(results5, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Infected, Day 1 vs. Infected, Day 3") +
  ylab("-Log10(Adjusted p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none",
        plot.title = element_text(size = 9))

tmp6 <- topTreat(tfit, coef = 6, n = Inf)
results6 <- mutate(tmp6, sig=ifelse(tmp6$adj.P.Val<0.1, "Sig", "Not Sig"))
p6 <- ggplot(results6, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Infected, Day 3 vs. Infected, Day 5") +
  ylab("-Log10(Adjusted p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none",
        plot.title = element_text(size = 9))


tmp7 <- topTreat(tfit, coef = 7, n = Inf)
results7 <- mutate(tmp7, sig=ifelse(tmp7$adj.P.Val<0.1, "Sig", "Not Sig"))
p7 <- ggplot(results7, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Infected, Day 5 vs. Infected, Day 14") +
  ylab("-Log10(Adjusted p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none",
        plot.title = element_text(size = 9))

grid.arrange(p1, p2, p3, p4, p5, p6, p7, nrow = 3)
```

```{r message=FALSE, warning=FALSE}
tmp1a <- tmp1 %>% select(gene_id, adj.P.Val, logFC, sprot_Top_BLASTX_hit, Kegg) %>% separate(sprot_Top_BLASTX_hit, into = c("SwissProt_GeneName", NA, NA, NA, NA, "sprot2", NA), "\\^") %>%   separate(sprot2, sep = "=", into = c(NA, "SwissProt_GeneFunction")) %>%  separate(Kegg, into = c("KEGG_ID", "KO_ID"), sep = "\\`") %>%  mutate(comp = "CtlvI1")
tmp2a <- tmp2 %>% select(gene_id, adj.P.Val, logFC, sprot_Top_BLASTX_hit, Kegg) %>% separate(sprot_Top_BLASTX_hit, into = c("SwissProt_GeneName", NA, NA, NA, NA, "sprot2", NA), "\\^") %>%   separate(sprot2, sep = "=", into = c(NA, "SwissProt_GeneFunction")) %>%  separate(Kegg, into = c("KEGG_ID", "KO_ID"), sep = "\\`") %>% mutate(comp = "CtlvI3")
tmp3a <- tmp3 %>% select(gene_id, adj.P.Val, logFC, sprot_Top_BLASTX_hit, Kegg) %>% separate(sprot_Top_BLASTX_hit, into = c("SwissProt_GeneName", NA, NA, NA, NA, "sprot2", NA), "\\^") %>%   separate(sprot2, sep = "=", into = c(NA, "SwissProt_GeneFunction")) %>%  separate(Kegg, into = c("KEGG_ID", "KO_ID"), sep = "\\`") %>% mutate(comp = "CtlvI5")
tmp4a <- tmp4 %>% select(gene_id, adj.P.Val, logFC, sprot_Top_BLASTX_hit, Kegg) %>% separate(sprot_Top_BLASTX_hit, into = c("SwissProt_GeneName", NA, NA, NA, NA, "sprot2", NA), "\\^") %>%   separate(sprot2, sep = "=", into = c(NA, "SwissProt_GeneFunction")) %>%  separate(Kegg, into = c("KEGG_ID", "KO_ID"), sep = "\\`") %>% mutate(comp = "CtlvI14")
tmp5a <- tmp5 %>% select(gene_id, adj.P.Val, logFC, sprot_Top_BLASTX_hit, Kegg) %>% separate(sprot_Top_BLASTX_hit, into = c("SwissProt_GeneName", NA, NA, NA, NA, "sprot2", NA), "\\^") %>%   separate(sprot2, sep = "=", into = c(NA, "SwissProt_GeneFunction")) %>%  separate(Kegg, into = c("KEGG_ID", "KO_ID"), sep = "\\`") %>% mutate(comp = "I1vI3")
tmp6a <- tmp6 %>% select(gene_id, adj.P.Val, logFC, sprot_Top_BLASTX_hit, Kegg) %>% separate(sprot_Top_BLASTX_hit, into = c("SwissProt_GeneName", NA, NA, NA, NA, "sprot2", NA), "\\^") %>%   separate(sprot2, sep = "=", into = c(NA, "SwissProt_GeneFunction")) %>%  separate(Kegg, into = c("KEGG_ID", "KO_ID"), sep = "\\`") %>% mutate(comp = "I3vI5")
tmp7a <- tmp7 %>% select(gene_id, adj.P.Val, logFC, sprot_Top_BLASTX_hit, Kegg) %>% separate(sprot_Top_BLASTX_hit, into = c("SwissProt_GeneName", NA, NA, NA, NA, "sprot2", NA), "\\^") %>%   separate(sprot2, sep = "=", into = c(NA, "SwissProt_GeneFunction")) %>%  separate(Kegg, into = c("KEGG_ID", "KO_ID"), sep = "\\`") %>% mutate(comp = "I5vI14")

bind_rows(tmp1a, tmp2a, tmp3a, tmp4a, tmp5a, tmp6a, tmp7a) %>% 
  select(-SwissProt_GeneFunction, -KEGG_ID, -KO_ID) %>% 
  filter(gene_id %in% dt.tib$gene) %>% 
  mutate(logFoldCount = ifelse(adj.P.Val < 0.10, round(logFC, 2), "ns")) %>% 
  select(-adj.P.Val, -logFC) %>% 
  pivot_wider(names_from = comp, values_from = logFoldCount) %>% 
  kbl(booktabs = T, 
      longtable =T,
      caption ="Annotations for differentially expressed genes. ns denotes non-significant genes for each comparison and numerical values are the log(fold change) difference") %>%
  kable_styling(latex_options = c("repeat_header"))
```
